The Leaky Semicolon

### 2021-10-27 paper/intro.tex
done with section 2

For single-threaded programs, memory can be thought of as you might
expect: programs write to, and read from, memory references.
This can be thought of as a total order over memory actions ($\xwki$),
where each read has a matching \emph{fulfilling} write ($\xrf$),
for example:
\begin{gather*}
  \THREAD{\PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW{y}{2}\SEMI
    \PR{y}{r}\SEMI \PR{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{wx0}{\DW{x}{0}}{}
      \event{wx1}{\DW{x}{1}}{right=of wx0}
      \event{wy2}{\DW{y}{2}}{right=of wx1}
      \event{ry2}{\DR{y}{2}}{right=of wy2}
      \event{rx1}{\DR{x}{1}}{right=of ry2}
      \rf[out=15,in=165]{wy2}{ry2}
      \rf[out=15,in=165]{wx1}{rx1}
      \wki{wx0}{wx1}
      \wki{wx1}{wy2}
      \wki{wy2}{ry2}
      \wki{ry2}{rx1}
    \end{tikzinline}}
\end{gather*}

@@

This model extends naturally to the case of shared-memory concurrency, leading to a \emph{sequentially consistent}
semantics \cite{Lamport:1979:MMC:1311099.1311750}, in which \emph{program order} inside a thread implies
a total \emph{causal order} between read and write events, for example
(where $\SEMI\!$ has higher precedence than $\PAR$):
\begin{gather*}
  \THREAD{\PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW{y}{2}}
  \PAR
  \THREAD{\PR{y}{r}\SEMI \PR{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{wx0}{\DW{x}{0}}{}
      \event{wx1}{\DW{x}{1}}{right=of wx0}
      \event{wy2}{\DW{y}{2}}{right=of wx1}
      \event{ry2}{\DR{y}{2}}{right=3em of wy2}
      \event{rx1}{\DR{x}{1}}{right=of ry2}
      \rf[out=15,in=165]{wy2}{ry2}
      \rf[out=15,in=165]{wx1}{rx1}
      \wki{wx0}{wx1}
      \wki{wx1}{wy2}
      \wki{ry2}{rx1}
    \end{tikzinline}}
\end{gather*}
We can represent such an execution as a labelled partial order, or
\emph{pomset} \cite{DBLP:conf/lop/Pratt85,GISCHER1988199}.  A program may give rise to many
executions, each reflecting a different interleaving of the threads.

@@

This has led to \emph{relaxed memory models}, in which the requirement of
sequential consistency is weakened to only apply \emph{per-location}. This allows executions that
are inconsistent with program order, such as the following, which contains an
\emph{antidependency} $(\xwk)$:
\begin{gather*}
  \THREAD{\PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW{y}{2}}
  \PAR
  \THREAD{\PR{y}{r}\SEMI \PR{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{wx0}{\DW{x}{0}}{}
      \event{wx1}{\DW{x}{1}}{right=of wx0}
      \event{wy2}{\DW{y}{2}}{right=of wx1}
      \event{ry2}{\DR{y}{2}}{right=3em of wy2}
      \event{rx0}{\DR{x}{0}}{right=of ry2}
      \rf[out=15,in=165]{wy2}{ry2}
      \rf[out=15,in=165]{wx0}{rx0}
      \wki{wx0}{wx1}
      \wk[out=-165,in=-15]{rx0}{wx1}
    \end{tikzinline}}
\end{gather*}
% Reads must be ordered between blocking writes, creating
% \emph{antidependencies} such as $\DRP{x}{0}\xwk\DWP{x}{1}$.

In such models, the causal order between events is important, and includes
control and data dependencies ($\xpo$) to avoid paradoxical ``out of thin
air'' examples such as the following.  (We routinely elide initializing
writes when they are uninteresting.)
\begin{gather*}
  \THREAD{\PR{x}{r}\SEMI \IF r \THEN \PW{y}{1} \FI}
  \PAR
  \THREAD{\PR{y}{s}\SEMI \PW{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{rx1}{\DR{x}{1}}{}
      \event{wy1}{\DW{y}{1}}{right=of rx1}
      \event{ry1}{\DR{y}{1}}{right=3em of wy1}
      \event{wx1}{\DW{x}{1}}{right=of ry1}
      \rf[out=15,in=165]{wy1}{ry1}
      \rf[out=-165,in=-15]{wx1}{rx1}
      \po{rx1}{wy1}
      \po{ry1}{wx1}
    \end{tikzinline}}
\end{gather*}
This candidate execution forms a cycle in causal order, so is disallowed,
but this depends crucially on the control dependency
from $(\DR{x}{1})$ to $(\DW{y}{1})$, and the data dependency
from $(\DR{y}{1})$ to $(\DW{x}{1})$. If either is missing, then this execution
is acyclic and hence allowed. For example dropping the control dependency
results in:
\begin{gather*}
  \THREAD{\PR{x}{r}\SEMI \PW{y}{1}}
  \PAR
  \THREAD{\PR{y}{s}\SEMI \PW{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{rx1}{\DR{x}{1}}{}
      \event{wy1}{\DW{y}{1}}{right=of rx1}
      \event{ry1}{\DR{y}{1}}{right=3em of wy1}
      \event{wx1}{\DW{x}{1}}{right=of ry1}
      \rf[out=15,in=165]{wy1}{ry1}
      \rf[out=-165,in=-15]{wx1}{rx1}
      \po{ry1}{wx1}
    \end{tikzinline}}
\end{gather*}

While syntactic dependency calculation suffices for hardware models, it is
not preserved by common compiler optimizations. For example, if we calculate
control dependencies syntactically, then there is a dependency from
$(\DR{x}{1})$ to $(\DW{y}{1})$, and therefore a cycle in, the candidate
execution:
\begin{gather*}
  \THREAD{\PR{x}{r}\SEMI \IF r \THEN \PW{y}{1} \ELSE \PW{y}{1} \FI}
  \PAR
  \THREAD{\PR{y}{s}\SEMI \PW{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{rx1}{\DR{x}{1}}{}
      \event{wy1}{\DW{y}{1}}{right=of rx1}
      \event{ry1}{\DR{y}{1}}{right=3em of wy1}
      \event{wx1}{\DW{x}{1}}{right=of ry1}
      \rf[out=15,in=165]{wy1}{ry1}
      \rf[out=-165,in=-15]{wx1}{rx1}
      \po{rx1}{wy1}
      \po{ry1}{wx1}
    \end{tikzinline}}
\end{gather*}
A compiler may lift the assignment $\PW{y}{1}$ out of the conditional,
thus removing the dependency.

To address this, \citet{DBLP:journals/pacmpl/JagadeesanJR20} introduced
\emph{Pomsets with Preconditions} (\PwP{}), where events are labeled with logical
formulae.  Nontrivial preconditions are introduced by store actions (modeling
data dependencies) and conditionals (modeling control dependencies):
\begin{gather*}
  \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
  \\
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \bigmid \DW{z}{0}}{}
    \end{tikzinline}}
\end{gather*}
In this diagram, $(s {<} 1)$ is a control dependency and $(r{*}s){=}0$ is a
data dependency. Preconditions are updated as events are prepended (we assume the
usual precedence for logical operators): 
\begin{gather*}
  \tag{$\dagger$}\label{eq1}
  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
  \\
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{rx0}{\DR{x}{0}}{}
      \event{ry0}{\DR{y}{0}}{right=of rx0}
      \event{wz0}{(0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \bigmid \DW{z}{0}}{right=of ry0}
      \po{ry0}{wz0}
    \end{tikzinline}}
\end{gather*}
In this diagram there are two reads.  As evidenced by the arrow, the read of
$y$ is ordered before the write, reflecting possible dependency; the read of
$x$ is not, reflecting independency.  The dependent read of $y$ allows the
precondition of the write to weaken: now the old precondition need only be
satisfied assuming the hypothesis $(0{=}s)$.  The independent read of $x$
allows no such weakening.  Nonetheless, the precondition of the write is now
a tautology, and so can be elided in the diagram:
\begin{gather*}
  % \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
  % \\
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{rx0}{\DR{x}{0}}{}
      \event{ry0}{\DR{y}{0}}{right=of rx0}
      \event{wz0}{\DW{z}{0}}{right=of ry0}
      \po{ry0}{wz0}
    \end{tikzinline}}
\end{gather*}
In order for a pomset to be \emph{complete}, all preconditions must be
tautologies.  Thus, an execution of \eqref{eq1} that reads $\DR{y}{2}$ must
elide the write action to be considered complete.  For this reason, we must
allow read and write to generate empty pomsets, in addition to singletons.

@@

For example, we can recover the expected semantics
for \eqref{eq1} by choosing
the predicate transformer which is independent of $(\DR x0)$
but dependent on $(\DR y0)$, which is the transformer
which maps $\bForm$ to $(0{=}s) \limplies \bForm$.
(In subsequent diagrams, we only show predicate transformers for reads.)
\begin{gather*}
  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
  \\
  \hbox{\begin{tikzinline}[node distance=2ex and 1.2em]
      \xform{wz0i}{\bForm}{}
      \xform{rx0d}{(0{=}r) \limplies \bForm}{right=of wz0i}
      \xform{rx0ry0d}{(0{=}r) \limplies(0{=}s) \limplies \bForm}{right=of rx0d}
      \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of rx0ry0d}
      \event{rx0}{\DR{x}{0}}{above left=of rx0ry0d}
      \event{ry0}{\DR{y}{0}}{above right=of rx0ry0d}
      \event{wz0}{(0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \bigmid \DW{z}{0}}{right=of ry0}
      \po{ry0}{wz0}
      \xo{rx0}{rx0d}
      \xo{ry0}{ry0d}
      \xo{rx0}{rx0ry0d}
      \xo{ry0}{rx0ry0d}
    \end{tikzinline}}
\end{gather*}
In the diagram, the dotted lines indicate set inclusion into the index of the transformer-family.
As a quick correctness test, we can see that sequential composition is
associative in this case, since it does not matter whether we
associate to the left, with the intermediate step eliding $\DWP{z}{0}$ in the
diagram above,
% \begin{gather*}
%   \PR{x}{r}\SEMI \PR{y}{s}
%   \\
%   \nonumber
%   \hbox{\begin{tikzinline}[node distance=2ex and 1.5em]
%       \xform{rx0d}{(0{=}r) \limplies \bForm}{}
%       \event{rx0}{\DR{x}{0}}{right=of rx0d}
%       \xform{rx0ry0d}{(0{=}r) \limplies(0{=}s) \limplies \bForm}{right=of rx0}
%       \event{ry0}{\DR{y}{0}}{right=of rx0ry0d}
%       \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of ry0}
%       \xform{wz0i}{\bForm}{left=of rx0d}
%       \xo{rx0}{rx0d}
%       \xo{ry0}{ry0d}
%       \xo{rx0}{rx0ry0d}
%       \xo{ry0}{rx0ry0d}
%     \end{tikzinline}}
% \end{gather*}
or to the right, with the intermediate step:
\begin{gather*}
  \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
  \\
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
      \event{ry0}{\DR{y}{0}}{}
      \event{wz0}{(0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \bigmid \DW{z}{0}}{right=of ry0}
      \xform{ry0d}{(0{=}s) \limplies \bForm}{left=of ry0}
      \xform{wz0i}{\bForm}{left=of ry0d}
      \po{ry0}{wz0}
      \xo{ry0}{ry0d}
    \end{tikzinline}}
\end{gather*}
This is an instance of the general result that sequential composition forms a monoid.

>>>>

For single-threaded programs, memory can be thought of as you might
expect: programs write to, and read from, memory references.
This can be thought of as a total order over memory actions ($\xwki$),
where each read has a matching \emph{fulfilling} write ($\xrf$),
for example:
\begin{gather*}
  \THREAD{\PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW{y}{2}\SEMI
    \PR{y}{r}\SEMI \PR{x}{s}}
  \\[-.4ex]
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{wx0}{\DW{x}{0}}{}
      \event{wx1}{\DW{x}{1}}{right=of wx0}
      \event{wy2}{\DW{y}{2}}{right=of wx1}
      \event{ry2}{\DR{y}{2}}{right=of wy2}
      \event{rx1}{\DR{x}{1}}{right=of ry2}
      \rf[out=15,in=165]{wy2}{ry2}
      \rf[out=15,in=165]{wx1}{rx1}
      \wki{wx0}{wx1}
      \wki{wx1}{wy2}
      \wki{wy2}{ry2}
      \wki{ry2}{rx1}
    \end{tikzinline}}
\end{gather*}

@@

This model extends naturally to the case of shared-memory concurrency, leading to a \emph{sequentially consistent}
semantics \cite{Lamport:1979:MMC:1311099.1311750}, in which \emph{program order} inside a thread implies
a total \emph{causal order} between read and write events, for example
(where $\SEMI\!$ has higher precedence than $\PAR$):
\begin{gather*}
  \THREAD{\PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW{y}{2}}
  \PAR
  \THREAD{\PR{y}{r}\SEMI \PR{x}{s}}
  \\[-.4ex]
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{wx0}{\DW{x}{0}}{}
      \event{wx1}{\DW{x}{1}}{right=of wx0}
      \event{wy2}{\DW{y}{2}}{right=of wx1}
      \event{ry2}{\DR{y}{2}}{right=3em of wy2}
      \event{rx1}{\DR{x}{1}}{right=of ry2}
      \rf[out=15,in=165]{wy2}{ry2}
      \rf[out=15,in=165]{wx1}{rx1}
      \wki{wx0}{wx1}
      \wki{wx1}{wy2}
      \wki{ry2}{rx1}
    \end{tikzinline}}
\end{gather*}
We can represent such an execution as a labelled partial order, or
\emph{pomset} \cite{DBLP:conf/lop/Pratt85,GISCHER1988199}.  A program may give rise to many
executions, each reflecting a different interleaving of the threads.

@@

This has led to \emph{relaxed memory models}, in which the requirement of
sequential consistency is weakened to only apply \emph{per-location}. This allows executions that
are inconsistent with program order, such as the following, which contains an
\emph{antidependency} $(\xwk)$:
\begin{gather*}
  \THREAD{\PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW{y}{2}}
  \PAR
  \THREAD{\PR{y}{r}\SEMI \PR{x}{s}}
  \\[-.4ex]
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{wx0}{\DW{x}{0}}{}
      \event{wx1}{\DW{x}{1}}{right=of wx0}
      \event{wy2}{\DW{y}{2}}{right=of wx1}
      \event{ry2}{\DR{y}{2}}{right=3em of wy2}
      \event{rx0}{\DR{x}{0}}{right=of ry2}
      \rf[out=15,in=165]{wy2}{ry2}
      \rf[out=15,in=165]{wx0}{rx0}
      \wki{wx0}{wx1}
      \wk[out=-165,in=-15]{rx0}{wx1}
    \end{tikzinline}}
\end{gather*}
% Reads must be ordered between blocking writes, creating
% \emph{antidependencies} such as $\DRP{x}{0}\xwk\DWP{x}{1}$.

In such models, the causal order between events is important, and includes
control and data dependencies ($\xpo$) to avoid paradoxical ``out of thin
air'' examples such as the following.  (We routinely elide initializing
writes when they are uninteresting.)
\begin{gather*}
  \THREAD{\PR{x}{r}\SEMI \IF r \THEN \PW{y}{1} \FI}
  \PAR
  \THREAD{\PR{y}{s}\SEMI \PW{x}{s}}
  \\[-.4ex]
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{rx1}{\DR{x}{1}}{}
      \event{wy1}{\DW{y}{1}}{right=of rx1}
      \event{ry1}{\DR{y}{1}}{right=3em of wy1}
      \event{wx1}{\DW{x}{1}}{right=of ry1}
      \rf[out=15,in=165]{wy1}{ry1}
      \rf[out=-165,in=-15]{wx1}{rx1}
      \po{rx1}{wy1}
      \po{ry1}{wx1}
    \end{tikzinline}}
\end{gather*}
This candidate execution forms a cycle in causal order, so is disallowed,
but this depends crucially on the control dependency
from $(\DR{x}{1})$ to $(\DW{y}{1})$, and the data dependency
from $(\DR{y}{1})$ to $(\DW{x}{1})$. If either is missing, then this execution
is acyclic and hence allowed. For example dropping the control dependency
results in:
\begin{gather*}
  \THREAD{\PR{x}{r}\SEMI \PW{y}{1}}
  \PAR
  \THREAD{\PR{y}{s}\SEMI \PW{x}{s}}
  \\[-.4ex]
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{rx1}{\DR{x}{1}}{}
      \event{wy1}{\DW{y}{1}}{right=of rx1}
      \event{ry1}{\DR{y}{1}}{right=3em of wy1}
      \event{wx1}{\DW{x}{1}}{right=of ry1}
      \rf[out=15,in=165]{wy1}{ry1}
      \rf[out=-165,in=-15]{wx1}{rx1}
      \po{ry1}{wx1}
    \end{tikzinline}}
\end{gather*}

While syntactic dependency calculation suffices for hardware models, it is
not preserved by common compiler optimizations. For example, if we calculate
control dependencies syntactically, then there is a dependency from
$(\DR{x}{1})$ to $(\DW{y}{1})$, and therefore a cycle in, the candidate
execution:
\begin{gather*}
  \THREAD{\PR{x}{r}\SEMI \IF r \THEN \PW{y}{1} \ELSE \PW{y}{1} \FI}
  \PAR
  \THREAD{\PR{y}{s}\SEMI \PW{x}{s}}
  \\[-.4ex]
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{rx1}{\DR{x}{1}}{}
      \event{wy1}{\DW{y}{1}}{right=of rx1}
      \event{ry1}{\DR{y}{1}}{right=3em of wy1}
      \event{wx1}{\DW{x}{1}}{right=of ry1}
      \rf[out=15,in=165]{wy1}{ry1}
      \rf[out=-165,in=-15]{wx1}{rx1}
      \po{rx1}{wy1}
      \po{ry1}{wx1}
    \end{tikzinline}}
\end{gather*}
A compiler may lift the assignment $\PW{y}{1}$ out of the conditional,
thus removing the dependency.

To address this, \citet{DBLP:journals/pacmpl/JagadeesanJR20} introduced
\emph{Pomsets with Preconditions} (\PwP{}), where events are labeled with logical
formulae.  Nontrivial preconditions are introduced by store actions (modeling
data dependencies) and conditionals (modeling control dependencies):
\begin{gather*}
  \IF{s{>}0} \THEN \PW{z}{r{*}(s{-}1)} \FI
  \\
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{wz0}{(s {>} 0) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{}
    \end{tikzinline}}
\end{gather*}
In this diagram, $(s {>} 0)$ is a control dependency and $(r{*}(s{-}1)){=}0$ is a
data dependency. Preconditions are updated as events are prepended (we assume the
usual precedence for logical operators): 
\begin{gather*}
  %\tag{$\dagger$}\label{eq1}
  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{>}0} \THEN \PW{z}{r{*}(s{-}1)} \FI
  \\
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{rx0}{\DR{x}{1}}{}
      \event{ry0}{\DR{y}{1}}{right=of rx0}
      \event{wz0}{(1{=}s) \limplies (s{>}0) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{right=of ry0}
      \po{ry0}{wz0}
    \end{tikzinline}}
\end{gather*}
In this diagram there are two reads.  As evidenced by the arrow, the read of
$y$ is ordered before the write, reflecting possible dependency; the read of
$x$ is not, reflecting independency.  The dependent read of $y$ allows the
precondition of the write to weaken: now the old precondition need only be
satisfied assuming the hypothesis $(1{=}s)$.  The independent read of $x$
allows no such weakening.  Nonetheless, the precondition of the write is now
a tautology, and so can be elided in the diagram.

We can complete the execution by adding the required writes:
\begin{gather*}
  \PW{x}{1}
  \SEMI
  \PW{y}{1}
  \PAR
  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{>}0} \THEN \PW{z}{r{*}(s{-}1)} \FI
  \\
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{wx}{\DW{x}{1}}{}
      \event{wy}{\DW{y}{1}}{right=of wx}
      \event{rx}{\DR{x}{1}}{right=3em of wy}
      \event{ry}{\DR{y}{1}}{right=of rx}
      \event{wz}{\DW{z}{0}}{right=of ry}
      \po{ry}{wz}
      \rf[out=10,in=170]{wx}{rx}
      \rf[out=10,in=170]{wy}{ry}
    \end{tikzinline}}
\end{gather*}
In order for a \PwP{} to be \emph{complete}, all preconditions must be
tautologies and all reads must be fulfilled by matching writes.  The first
requirement captures the sequential semantics.  The second requirement
captures the concurrent semantics.  These correspond to two views of memory
for each thread: thread-local and global.  In a
\emph{multicopy-atomic} (\mca{}) architecture, there is only one global view,
shared by all processors, which is neatly captured by the order of the
pomset.

An untaken conditional produces no events.  \PwP{} models this by including
the empty pomset in the semantics of every program fragment.  To then ensure
that $\SKIP$ is not a refinement of $\PW{x}{1}$, \PwP{} include the
\emph{termination} action, $\aTerm{}$, which we have elided in the examples
above.

@@

For example, we can recover the expected semantics
for \eqref{eq1} by choosing
the predicate transformer which is independent of $(\DR x0)$
but dependent on $(\DR y0)$, which is the transformer
which maps $\bForm$ to $(0{=}s) \limplies \bForm$.
(In subsequent diagrams, we only show predicate transformers for reads.)
\begin{gather*}
  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
  \\
  \hbox{\begin{tikzinline}[node distance=2ex and 1.2em]
      \xform{wz0i}{\bForm}{}
      \xform{rx0d}{(0{=}r) \limplies \bForm}{right=of wz0i}
      \xform{rx0ry0d}{(0{=}r) \limplies(0{=}s) \limplies \bForm}{right=of rx0d}
      \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of rx0ry0d}
      \event{rx0}{\DR{x}{0}}{above left=of rx0ry0d}
      \event{ry0}{\DR{y}{0}}{above right=of rx0ry0d}
      \event{wz0}{(0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \bigmid \DW{z}{0}}{right=of ry0}
      \po{ry0}{wz0}
      \xo{rx0}{rx0d}
      \xo{ry0}{ry0d}
      \xo{rx0}{rx0ry0d}
      \xo{ry0}{rx0ry0d}
    \end{tikzinline}}
\end{gather*}
In the diagram, the dotted lines indicate set inclusion into the index of the transformer-family.
As a quick correctness test, we can see that sequential composition is
associative in this case, since it does not matter whether we
associate to the left, with the intermediate step eliding $\DWP{z}{0}$ in the
diagram above,
% \begin{gather*}
%   \PR{x}{r}\SEMI \PR{y}{s}
%   \\
%   \hbox{\begin{tikzinline}[node distance=2ex and 1.5em]
%       \xform{rx0d}{(0{=}r) \limplies \bForm}{}
%       \event{rx0}{\DR{x}{0}}{right=of rx0d}
%       \xform{rx0ry0d}{(0{=}r) \limplies(0{=}s) \limplies \bForm}{right=of rx0}
%       \event{ry0}{\DR{y}{0}}{right=of rx0ry0d}
%       \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of ry0}
%       \xform{wz0i}{\bForm}{left=of rx0d}
%       \xo{rx0}{rx0d}
%       \xo{ry0}{ry0d}
%       \xo{rx0}{rx0ry0d}
%       \xo{ry0}{rx0ry0d}
%     \end{tikzinline}}
% \end{gather*}
or to the right, with the intermediate step:
\begin{gather*}
  \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
  \\
  \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
      \event{ry0}{\DR{y}{0}}{}
      \event{wz0}{(0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \bigmid \DW{z}{0}}{right=of ry0}
      \xform{ry0d}{(0{=}s) \limplies \bForm}{left=of ry0}
      \xform{wz0i}{\bForm}{left=of ry0d}
      \po{ry0}{wz0}
      \xo{ry0}{ry0d}
    \end{tikzinline}}
\end{gather*}
This is an instance of the general result that sequential composition forms a monoid.
